Supporting information: Copulas and their potential for ecology

### 2019-09-14 Paper.Rmd
background section complete (though still may need some editing)

# Abstract
<!--***DANr: I have not edited this abstract at all since Ecol Monogr-->
All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also help reveal causal relationships between variables. We also describe consequences of copula structure for ecological phenomena, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology, benefiting the discipline.

@@

<!--DONE ***DANr: reword this para so people don't think the ranks *are* the 
copulas, and to try to pre-defend against some of what ref 1 said. Also,
state that these are conceptual introductions only and are not formal,
a more comprehensive introduction is in xyz-->
Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect tail dependence and asymmetries of tail dependence. 
Tail dependence (defined precisely in the section on Background, below) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below). Although copulas can be used to model the entire
complex dependence structure of data, we have found tail dependence to be
a useful aspect of the full structure of dependence, so tail dependence 
is one of the main foci of this paper.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
second through
fourth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (non-N) copula structure that
were examined (in the fifth and sixth sections of the text). Lower boxes 
correspond to potential consequences we examined of non-N copula structure for
ecological understanding and for applications (in the seventh and eighth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

<!--DONE ***DANr: Maybe use the words tail dependence more and the word copula
less in the below few paras? Generally the words "copula structure" may be rubbing some
people wrong. Replacement words? What do you actually mean here? Maybe call it
asymmetric dependence, or complex dependence structure?-->
One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum. Liebig's law is 
the idea that growth is controlled not by total resources but by the resources which
are scarcest relative to organism needs. 
If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association are visible in a rank-based
plot, but may not be revealed by standard correlation techniques.

A second reason we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated (using copulas) complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail dependence between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

<!--DONE ***DANr: Need some caveating here to accommodate the people who slavishly
follow the old maxim that correlation is not causation.-->
A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Although other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from that standard or default cases. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and tail dependence in ecology? And (Q3) what are the 
consequences of non-normal copula
structure and tail dependence for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have been used very effectively, but rarely so far in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

Thus Sklar's theorem implies that
any random vector can be constructed from a copula and marginal 
distributions. And conversely,
any copula and any univariate distribution functions give rise to a random vector. 
Marginals contain no information about the
dependence structure of a random vector, so the copula contains
all such information - it is a complete description of the dependence
between variables. Sklar's theorem makes it possible to construct
bivariate distributions in two separate steps: by specifying the marginal 
distributions and by specifying the dependence structure. For instance,
construction of several bivariate distributions is carried out in this way in
Fig. \ref{PedagogFigCopThry}. The contrast between Fig. \ref{PedagogFigCopThry}C and D
illustrates in particular how a bivariate distribution can be changed, while retaining 
the same marginals, by changing the copula. Fig. \ref{PedagogFigCopThry}C is a 
bivariate normal distribution, but Fig. \ref{PedagogFigCopThry}D clearly 
is not, although its marginals
are the same. Fig. \ref{PedagogFigCopThry}D shows stronger left- than right-tail dependence. 
Sklar's theorem
also makes it possible to study and model dependence separately 
from marginal distributions. For instance, in a pedagogical 
example, @Anderson2018 modelled the dependence between abundance
data for two fish species, red moki and black angelfish, by first 
modelling the marginal distributions and then modelling 
the dependence using Gaussian copulas (their section 2).
See the Discussion for a few words on multivariate copulas, which can
be developed in the same way as above [@nelsen2006_copula; @joe2014_dependence;
@MaiScherer2017; @Anderson2018].
<!--***DAN: dont forget to put these few words in!-->

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family (henceforth abbreviated N), which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of dependence
between the variables. An N copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs correspondng to bivariate N copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
tail dependence (see
below for a more precise defintion of tail dependence). 
The Clayton copula family (henceforth abbreviated C),
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to N copulas, C copulas
have stronger left- than right-tail dependence. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). The survival
Clayton family (here abbreviated SC) is the symmetric opposite of the C family, showing
stronger right- than left tail dependence (Fig. \ref{SM-figped_C}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail dependence, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

Having introduced tail dependence in a conceptual way and referred conceptually to it
in examples, we now need a more precise definition. 
<!--***DAN: fill in-->

@@

<!--***DANr: Add a new section called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
1) DONE - The definition, Sklar's theorem
2) DONE What good are copulas - intuitive idea, basically, but also have to be linked to 
the formalities of the definition
3) DONE What some of the different copula families look like - in particular you will use BB1 in Fig S2 and cite Background, so mention it
  Three new figures: Normal, Clayton, BB1
STILL HAVE TO DO 4) What tail dependence actually is
DONE 5) The explanation of how ranks relate to copulas and estimating the copula parameters
How to carry out fitting (basically, MLE fitting on ranks - Genest et al paper)
-->

@@

<!--***DAN: text storage for later:
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. -->

>>>>

# Abstract
<!--***DANr: I have not edited this abstract at all since Ecol Monogr, it needs to be redone-->
All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also help reveal causal relationships between variables. We also describe consequences of copula structure for ecological phenomena, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology, benefiting the discipline.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*; copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cummulative distribution function (cdf) of the $x_i$ and $y_i$,
respectively. See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals. Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables; 
they provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the section on Background, below) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail association 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below). Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to be
a useful aspect of the full structure of dependence, so tail associations 
is one of the main foci of this paper.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
third through
fifth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (non-N) copula structure that
were examined (in the sixth and seventh sections of the text). Lower boxes 
correspond to potential consequences we examined of non-N copula structure for
ecological understanding and for applications (in the eighth and ninth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

One way we may expect, *a priori*, a copula 
study of complex dependence structures between
ecological variables
to better reflect ecological relationships relates to Liebig's law of the
minimum. Liebig's law is 
the idea that growth is controlled not by total resources but by the resources which
are scarcest relative to organism needs. 
If, for instance, the growth of a plant depends on soil 
nitrogen, N, and other factors, a plot of plant growth rates versus soil N may 
look more like Fig. \ref{fig_cop_pedag1}A or C than like 
Fig. \ref{fig_cop_pedag1}B or D. Nitrogen 
will control plant growth, producing a clear
relationship, only when it is limiting, in the left portions of the distributions. 
The portions of the distributions that dominate the association are visible in a rank-based
plot, but may not be revealed by standard correlation techniques.

A second reason we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated (using copulas) complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail associations between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Although other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also more strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand, stronger left-tail associations in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from that standard or default cases. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3) what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have been used very effectively, but rarely so far in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

@@

Thus Sklar's theorem implies that
any random vector can be constructed from a copula and marginal 
distributions. And conversely,
any copula and any univariate distribution functions give rise to a random vector. 
Marginals contain no information about the
dependence structure of a random vector, so the copula contains
all such information - it is a complete description of the dependence
between variables. Sklar's theorem makes it possible to construct
bivariate distributions in two separate steps: by specifying the marginal 
distributions and by specifying the dependence structure. For instance,
construction of several bivariate distributions is carried out in this way in
Fig. \ref{PedagogFigCopThry}. The contrast between Fig. \ref{PedagogFigCopThry}C and D
illustrates in particular how a bivariate distribution can be changed, while retaining 
the same marginals, by changing the copula. Fig. \ref{PedagogFigCopThry}C is a 
bivariate normal distribution, but Fig. \ref{PedagogFigCopThry}D clearly 
is not, although its marginals
are the same. Fig. \ref{PedagogFigCopThry}D shows stronger association of the two variables
in the left than in the right tails. Sklar's theorem
also makes it possible to study and model dependence separately 
from marginal distributions. For instance, in a pedagogical 
example, @Anderson2018 modelled the dependence between abundance
data for two fish species, red moki and black angelfish, by first 
modelling the marginal distributions and then modelling 
the dependence using Gaussian copulas (their section 2).
See the Discussion for a few words on multivariate copulas, which can
be developed in the same way as above [@nelsen2006_copula; @joe2014_dependence;
@MaiScherer2017; @Anderson2018].
<!--***DAN: dont forget to put these few words in!-->

@@

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family (henceforth abbreviated N), which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of dependence
between the variables. An N copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs correspondng to bivariate N copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails (see
below for a more precise defintion of tail associations). 
The Clayton copula family (henceforth abbreviated C),
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to N copulas, C copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). The survival
Clayton family (here abbreviated SC) is the symmetric opposite of the C family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_C}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

<!--***DAN: Note: I am using tail dependence for the specific thing defined on p 62 of Joe,
and tail associations as a more general term. I have changed this in parts of the ms
upsteam from this point, but need to remember to change it as I move downstream.-->

Having introduced tail association conceptually and referred to it
in examples, we now need a more precise definition of a measure of tail association.
We use the term *tail association* to describe the general idea of the strength of
associations between two variables in the tails of their distributions. One of the
ways this is measured is called *tail dependence* (@joe2014_dependence, section 2.13; 
@nelsen2006_copula, section 5.4), which we here define. Given a random vector $(X,Y)$
with margins $F_X$, $F_Y$ and defining $U=F_X(X)$, $V=F_Y(Y)$, the upper-tail
dependence of $X$ and $Y$ is defined as 
$\lambda_U = \lim_{u \rightarrow 1^-} P[Y>F_Y^{-1}(u) | X>F_X^{-1}(u)]$. This equals
$\lim_{u \rightarrow 1^-} P[U>u | V>u]$, which in turn equals 
$\lim_{u \rightarrow 1^-} P[U>u, V>u]/P[V>u]=\lim_{u \rightarrow 1^-} P[U>u, V>u]/(1-u)$,
which shows upper-tail dependence is defined symmetrically in the two random variables.
Lower tail dependence is defined analogously as 
$\lambda_L = \lim_{u \rightarrow 1^-} P[Y \leq F_Y^{-1}(u) | X \leq F_X^{-1}(u)]$.
Tail dependence is a property of the random variable $(U,V)$, so depends only
on the copula and not on the marginals combined with it. It can be shown 
(@nelsen2006_copula, section 5.4) that $\lambda_L=\lambda_U=0$ for the N copula,
$\lambda_L=2^{-1/p}$ and $\lambda_U=0$ for a C copula with parameter $p>0$,
and $\lambda_L=0$ and $\lambda_U=2^{-1/p}$ for an SC copula with parameter $p>0$.
In the section on methods for Q1 we will introduce
ways tail dependence can be applied to data, by first fitting copulas to the data. We will also 
introduce nonparametric measures of tail association that apply directly to data.

@@

@@

### 2019-09-14 SupportingInformation.Rmd
background section complete (though still may need some editing)

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption[Lower- and upper-tail dependence can vary independently of correlation]{Lower- and 
upper-tail dependence can vary independently of correlation. (A-C) Data generated using the BB1 
copula (see Background section in the main text) 
with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths, but with the same 
Kendall correlation (K) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column.\label{fig_cop_pedag}}
\end{center}
\end{figure}

>>>>

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm,height=8 cm]{./Results/copula_pedagog_fig_BB1.pdf}
\caption[Lower- and upper-tail dependence can vary independently of correlation]{Lower- and 
upper-tail dependence can vary independently of correlation. (A-C) Data generated using the BB1 
copula (see Background section in the main text) 
with three different 
sets of parameters chosen to produce a variety of 
lower-tail (LT) and upper-tail (UT) dependence strengths, but with the same 
Kendall correlation (K) (to two digits) in all cases. (D-F) Similar to A-C, but
with stronger correlation. Asymmetry of tail dependence (LT-UT) is the same 
within each column. See Background section in the main text for a formal description of 
tail dependence and other measures of association in the tails.\label{fig_cop_pedag}}
\end{center}
\end{figure}
